---
title: ""
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Mental Health Service Encounter Hours by Age, Gender Identity, and Race and Facility Technical Documentation

### **Group 1 – Akhil Kodali, Heather Davis, Scott Bui**

</br>


##  DATA CLEANING

### Data Sources

-   Heartland Family Services. (2021). HFS Service Data \[Data file\]. Retrieved from: https://unomaha.instructure.com/courses/50683/files/5256675?module_item_id=1562501

    The primary data source we used for our data cleaning was the HFS Service Data.csv file from the HFS resources section of Canvas. This file was provided to our class by Dr. Clayton Juarez, Ph.D.of Heartland Family Services (HFS) in Omaha, Nebraska. Dr.Juarez pulled the data from a system HFS uses to track the mental health services it provides to clients. The file contained 51 columns and 8746 rows. The primary source contained column headings such as: gender, program\_name, program\_type, and facility.

-   Heartland Family Services. (2021). SLA Data Columns 2021-08-25 \[Data file\]. Retrieved from: https://unomaha.instructure.com/courses/50683/files/5217061?module_item_id=1562500

    The secondary data source we used for our data cleaning was the SLA Data Columns 2021-08-25.xlsx file from the HFS resources section of Canvas. This file was also provided to us by Dr. Juarez. The file contains 3 columns and 52 rows. It provides the field names, notes, and descriptions about the fields. From the SLA Data Columns file, we were able to discern helpful information such as the meaning of the codes provided for race and duration. 

### Intellectual Policy Constraints

Permission for us to use this information was given to us by Heartland Services. Even though the information has been cleaned of personal information, we are still bound to observe data privacy standards.

### Metadata

The source we had for Metadata was the SLA Data Columns document described in the preceding Data Sources section. This document contained short descriptions of the fields, often just writing a longer description of the field name, but sometimes listing out the meanings of codes used in the columns.


### Data Issues

-   gender\_identity - In the gender identity column, we found both the value of female and woman and felt this was duplicate, especially since we did not see an equivalent with the value of man. Only man was used and not male. 

-   simple\_race - In the simple race column, we found values that did not correspond to the numeric codes provided in the SLA Data Columns document.

### Data Remediation

-   Removed rows we did not need for our research.
-   is\_approved - We only kept the rows from the data that had a value of 1 from this column. These were records that were approved by the supervisor.
-   simple\_race - We removed the rows in the data that did not correlate with the values provided in the SLA Data Columns
    document.We also replaced the numeric codes for race with their label values to make them easier to understand. For example, the value 2 was replaced with Alaskan Native.
-   gender\_identity - We changed the value of female to woman to correspond to the value of man because male was not used for individuals who identified themselves as men. We also changed the value to Not Obtained to NA since we felt they were very similar.


### Data Cleaning Code

**1. Loading the required R studio packages**

``` {r message = FALSE,warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(tinytex)
library(rticles)
```

**2. Loading the data from the CSV file**

``` {r}
Full_HFS_Data <- read.csv("HFS Service Data.csv")
```

**3. Selecting the columns that are required for our analysis**

``` {r}
HFS_1 <- select(Full_HFS_Data, c(program_name, facility, job_title,actual_date, event_name, is_approved, NormalWorkHours,duration_num, total_duration_num, recordID,state, age, simple_race, gender_identity))
```

**4. Selecting the data that is approved by the supervisor = 1**

``` {r}
HFS_2 <- subset(HFS_1, is_approved == 1) 
```

Here is\_approved = 1 means the entered records that were approved by the supervisor. So, only approved data are
considered as there is no point in analyzing unapproved data.

**5. Omitting the invalid simple\_race data**

``` {r}
HFS_Data <- subset(HFS_2,  simple_race != 21 & simple_race != 65 &simple_race != 5 & simple_race != 9 & simple_race != 17 & simple_race != 20  & simple_race != 81 )
```

The simple race having numbers 21, 65, 5, 9, 17, 20, 81 did not align with the simple race data key in the SLA. So, it was likely mis-keyed and we are removing the false data here.

**6. Replacing the numeric codes in simple\_race to actual words.**

``` {r}
HFS_Data$simple_race[HFS_Data$simple_race ==1 ] <- "Caucasian"
HFS_Data$simple_race[HFS_Data$simple_race ==2 ] <- "Alaskan Native"
HFS_Data$simple_race[HFS_Data$simple_race ==4 ] <- "American Indian/Native American"
HFS_Data$simple_race[HFS_Data$simple_race ==8 ] <- "Asian"
HFS_Data$simple_race[HFS_Data$simple_race ==16 ] <- "Black / African-American"
HFS_Data$simple_race[HFS_Data$simple_race ==32] <- "Hawaiian Native / Pacific Islander"
HFS_Data$simple_race[HFS_Data$simple_race ==64] <- "Two or more races"
HFS_Data$simple_race[HFS_Data$simple_race ==128 ] <- "Other"
HFS_Data$simple_race[HFS_Data$simple_race ==0 ] <- "Unknown or Not Collected"
```

We replaced the race code numbers with the exact words so that it will be easier to unsderstand as labels.

**7. Modifying the column gender\_identity**

``` {r}
HFS_Data$gender_identity[HFS_Data$gender_identity == "Female"] <- "Woman"
HFS_Data$gender_identity[HFS_Data$gender_identity == "Other"] <- "Non-Binary"
HFS_Data$gender_identity[HFS_Data$gender_identity == "Two-Spirit"] <- "Non-Binary"
```

We modified the values “Female” to “Woman” because they both mean the same. We chose "Woman" as the label because there is no corresponding "Male" label, only "Man". We have modified the values “Other” and “Two-Spirit” to “Non-Binary” as two spirit is both male and female spirit which is non-binary and also if they are not man or woman they are non-binary.

**8. Modifying the values containing “NA” in data to “Not Obtained”**

``` {r}
HFS_Data[is.na(HFS_Data)] <- "Not Obtained"
```

We combined "NA" and "Not Obtained" because they were very similar. 

**9. Modifying the actual\_date column**

``` {r warning=FALSE}
HFS_Data$actual_date <- as.numeric(HFS_Data$actual_date)
HFS_Data <- HFS_Data %>% mutate(actual_date = case_when(actual_date >= 3287  & actual_date <= 3652 ~ 'Year10',
                                                        actual_date >= 2923  & actual_date <= 3287 ~ 'Year9',
                                                        actual_date >= 2557  & actual_date <= 2922 ~ 'Year8',
                                                        actual_date >= 2192  & actual_date <= 2556 ~ 'Year7',
                                                        actual_date >= 1827  & actual_date <= 2191 ~ 'Year6',
                                                        actual_date >= 1462  & actual_date <= 1826 ~ 'Year5',
                                                        actual_date >= 1096  & actual_date <= 1461 ~ 'Year4',
                                                        actual_date >= 731  & actual_date <= 1095 ~ 'Year3',
                                                        actual_date >= 366  & actual_date <= 730 ~ 'Year2',
                                                        actual_date >= -7  & actual_date <= 365   ~ 'Year1'))
```

We filtered the days with respect to years from the beginning to the end so that evaluating the results is easier.


**10. Verifying the data has been modified according to our needs**

The following code checks the values for the simple race, gender identity, and is approved columns, as well as the column headers.

``` {r}
sort(unique(HFS_Data$simple_race))
```

    ## [1] "American Indian/Native American"    "Asian"                             
    ## [3] "Black / African-American"           "Caucasian"                         
    ## [5] "Hawaiian Native / Pacific Islander" "Other"                             
    ## [7] "Two or more races"                  "Unknown or Not Collected"

``` r
unique(HFS_Data$gender_identity)
```

    ## [1] "Not Obtained"            "Woman"                  
    ## [3] "Man"                     "Client Declined to Give"
    ## [5] "Non-Binary"

``` r
unique(HFS_Data$is_approved)
```

    ## [1] 1

**11. Viewing the modified data**

``` {r}
str(HFS_Data)
```

    ## 'data.frame':    7989 obs. of  14 variables:
    ##  $ program_name      : chr  "Mental Health" "Mental Health" "Mental Health" "Mental Health" ...
    ##  $ facility          : chr  "Center Mall Office" "Center Mall Office" "Center Mall Office" "Center Mall Office" ...
    ##  $ job_title         : chr  "THERAPIST II" "THERAPIST II" "THERAPIST II" "THERAPIST II" ...
    ##  $ actual_date       : chr  "Year3" "Year2" "Year2" "Year2" ...
    ##  $ event_name        : chr  "Collateral Note" "Individual Therapy" "Individual Therapy" "Individual Therapy" ...
    ##  $ is_approved       : int  1 1 1 1 1 1 1 1 1 1 ...
    ##  $ NormalWorkHours   : chr  "Yes" "Yes" "Yes" "Yes" ...
    ##  $ duration_num      : int  2 51 0 50 80 4 2 52 115 0 ...
    ##  $ total_duration_num: int  2 61 0 60 90 4 2 62 125 0 ...
    ##  $ recordID          : int  338 338 338 338 338 338 338 338 338 338 ...
    ##  $ state             : chr  "NE" "NE" "NE" "NE" ...
    ##  $ age               : int  26 25 25 25 25 25 26 25 25 25 ...
    ##  $ simple_race       : chr  "Black / African-American" "Black / African-American" "Black / African-American" "Black / African-American" ...
    ##  $ gender_identity   : chr  "Not Obtained" "Not Obtained" "Not Obtained" "Not Obtained" ...

 **12. Selecting only Mental Health from the three services provided**

``` {r}
nofvisits <- data.frame(count(HFS_Data,actual_date,program_name,facility))
ggplot(data = nofvisits,aes(x=actual_date,y=n,na.rm=TRUE))+geom_bar(stat="identity")+facet_wrap(~program_name)+ 
ggtitle("Total visits by years from the start of service for different program names")+
theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1)) 
```


We can observe from the graph above that visits to mental health programs are higher when compared to other programs. As a result, we chose to focus our research solely on mental health, ignoring substance use and gambling.

 **13. Viewing box plots of different programs with duration**

``` {r}
ggplot(HFS_Data, aes(program_name,duration_num,by1=recordID,fill=program_name)) + 
  geom_boxplot(aes(program_name), alpha=0.3) +
  xlab("Programs offered") +ylab("Duration")
```



``` {r}
  theme(legend.position = "none")
```

    ## List of 1
    ##  $ legend.position: chr "none"
    ##  - attr(*, "class")= chr [1:2] "theme" "gg"
    ##  - attr(*, "complete")= logi FALSE
    ##  - attr(*, "validate")= logi TRUE

Using a box plot of the three programs, we identified outliers in the mental health program. 

 **14. Removing the unwanted outliers**

``` {r}
HFS_Data <- HFS_Data %>%
  filter (duration_num < 350 & duration_num > 10)
```

We have removed the outliers which we encountered in the box plot above.

 **15. Additional Data cleaning for Facility**

``` {r clean group facility into categories}
HFS_Data$facility[HFS_Data$facility == "Center Mall Office"] <- "Center"
HFS_Data$facility[HFS_Data$facility == "North Omaha Intergenerational Campus (Service)"] <- "NOIC"
HFS_Data$facility[HFS_Data$facility == "Heartland Family Service - Logan"] <- "HFS-Other"
HFS_Data$facility[HFS_Data$facility == "Heartland Family Service - Gendler"] <- "HFS-Gendler"
HFS_Data$facility[HFS_Data$facility == "Heartland Family Service - Child and Family Center"] <- "HFS-Other"
HFS_Data$facility[HFS_Data$facility == "Heartland Family Service - Central"] <- "HFS-Other"
HFS_Data$facility[HFS_Data$facility == "Heartland Family Service - Family Works Nebraska"] <- "HFS-Other"
HFS_Data$facility[HFS_Data$facility == "Heartland Family Service - Sarpy"] <- "HFS-Other"
HFS_Data$facility[HFS_Data$facility == "Heartland Family Service - Glenwood"] <- "HFS-Other"
HFS_Data$facility[HFS_Data$facility == "Heartland Family Service - Heartland Homes"] <- "HFS-Other"
HFS_Data$facility[HFS_Data$facility == "Heartland Family Service - Lakin"] <- "HFS-Other"
HFS_Data$facility[HFS_Data$facility == "Omaha (Spring) Reporting Center"] <- "RepCntr"
HFS_Data$facility[HFS_Data$facility == "Omaha (Blondo) Reporting Center"] <- "RepCntr"
HFS_Data$facility[HFS_Data$facility == "Bellevue Reporting Center"] <- "RepCntr"
HFS_Data$facility[HFS_Data$facility == "Sanctuary House"] <- "SancHouse"
HFS_Data$facility[HFS_Data$facility == "Lewis Central High School"] <- "School"
HFS_Data$facility[HFS_Data$facility == "Lewis Central Middle School"] <- "School"
HFS_Data$facility[HFS_Data$facility == "Kirn Junior High  School"] <- "School"
HFS_Data$facility[HFS_Data$facility == "Wilson Middle School"] <- "School"
HFS_Data$facility[HFS_Data$facility == "Thomas Jefferson High School"] <- "School"
HFS_Data$facility[HFS_Data$facility == "Kreft Primary School"] <- "School"
HFS_Data$facility[HFS_Data$facility == "Abraham Lincoln High School"] <- "School"
HFS_Data$facility[HFS_Data$facility == "Titan Hill Intermediate School"] <- "School"
HFS_Data$facility[HFS_Data$facility == "Kanesville Alternative Learning Center"] <- "School"
```

We  reassigned the names of facilities for better analysis of data. We grouped all the facilities having school into School, reporting canter as RepCntr, HFS as HFS-Other and HFS\_Gendler. Here we separated  the HFS-Gendler from other HFS facilities because HFS\_Gendler had more hours associated with it than the other HFS locations.

 **16. Removing data from other states than Nebraska and Iowa**

``` {r}
HFS_Data <- subset(HFS_Data,program_name == 'Mental Health')
HFS_Data <- HFS_Data %>% filter(state=="NE" | state=="IA")
HFS_Data <- HFS_Data %>% mutate(ageg = case_when(age >= 55  & age <= 72 ~ '55-72',
                                                     age >= 37  & age <= 54 ~ '37-54',
                                                     age >= 19  & age <= 36 ~ '19-36',
                                                     age >= 0  & age <= 18 ~ '1-18'))
```

HFS facilities are located in Iowa and Nebraska and clients from from Iowa and Nebraska are the most frequent users of the HFS services, so we removed clients from other states.

## R SCRIPT AND RESULTS

The next three plots depict the average duration hours by age, gender identity, and race, by facility and state.  

#### Plot 1 – Mental Health Service Encounter Hours by Age and Facility

``` {r}
age<-aggregate(duration_num~ageg+facility+state,HFS_Data,FUN=mean)
age$duration_num<-round(age$duration_num,digits=1)

ggplot(age,aes(x=ageg,y=duration_num,na.rm=TRUE,fill=state))+geom_bar(stat="identity",position="dodge")+facet_wrap(~facility,ncol=2)+geom_text(aes(label=duration_num), position=position_dodge(width=1), vjust=1,size=3)+
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1)) +labs(y= "Average service hours spent", x = "Age Groups",title = "Service hours spent on different age groups by facility and state")
```


**Age Groups:** We can see from the above graph that the age group 37-54 has the longest average duration service encounter, followed by the age groups 19-36 and 55-72. Also, we notice that NOIC and School facilities are exclusively visited by people aged 1 to 18. Patients from both states are treated at the facilities Center and HFS-other. Clients of various ages come to Micah House from Iowa, but only those between the ages of 19 and 36 come from Nebraska. In those facilities that serve both Iowa and Nebraska older clients, we do not see a significant difference between the average duration of the service encounters. Overall, regardless of age group, Iowa clients appear to require more service time than those in Nebraska. The only significance we took away from the data is that the average service encounter time is more closely associated with facility type than age – with the biggest difference between schools and reporting centers, followed by schools and sanctuary houses.

#### Plot 2 – Mental Health Service Encounter Hours by Gender Identity and Facility

``` {r}
gender<-aggregate(duration_num~gender_identity+facility+state,HFS_Data,FUN=mean)
gender$duration_num<-round(gender$duration_num,digits=1)

ggplot(gender,aes(x=gender_identity,y=duration_num,na.rm=TRUE,fill=state))+geom_bar(stat="identity",position="dodge")+
  facet_wrap(~facility,ncol=2) +geom_text(aes(label=duration_num), position=position_dodge(width=1), vjust=1,size=3)+
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1)) +
  labs(y= "Average service hours spent", x = "Gender",title = "Service hours spent on different genders by facility and state")
```



**Gender Identity:** The most average service encounter hours are required by women from Nebraska’s reporting center facilities. If we look at the schools, the least number of average service encounter hours necessary for treatment is for men. We may also see that the schools only have clients from Iowa, whereas sancuary houses and reporting centers only have clients from Nebraska. If we look at all of the facilities, we can conclude that women require similar treatment hours than men. There is also a twenty-point difference between non-binary clients and women and even more significant difference between non-binary clients and men at the HFS-Gendler location. Non-binary clients did not represent a significant difference elsewhere, but they were not identified at all locations. Again, the most significant differences between clients seems to be based on facility type – with the biggest difference between schools and reporting centers, followed by schools and sanctuary houses.



#### Plot 3 – Mental Health Service Encounter Hours by Race and Facility

``` {r}
race<-aggregate(duration_num~simple_race+facility+state,HFS_Data,FUN=mean)
race <- race %>% filter(simple_race=="Asian" | simple_race=="American Indian/Native American" | 
                          simple_race=="Black / African-American" | simple_race=="Caucasian" | simple_race=="Hawaiian Native / Pacific Islander")
race$simple_race[race$simple_race == "American Indian/Native American"] <- "Native American"
race$simple_race[race$simple_race == "Hawaiian Native / Pacific Islander"] <- "Hawaiian Native"
race$duration_num<-round(race$duration_num,digits=1)

ggplot(race,aes(x=simple_race,y=duration_num,na.rm=TRUE,fill=state))+geom_bar(stat="identity",position="dodge")+
  facet_wrap(~facility,ncol=2)+geom_text(aes(label=duration_num), position=position_dodge(width=1), vjust=1,size=3)+
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1)) +
    labs(y= "Average service hours spent", x = "Race",title = "Service hours spent on different races by facility and state")
```



**Race:** If we compare all of the facilities, Native Americans have the longest average service encounter durations, as shown in the graph above. For the reporting centers, the most longest average service encounter durations are with Native Americans. Among Native American clients at the HRS-Other locations, Native Americans from Iowa receive more than double the average amount of service encounter hours than Nebraska Native Americans. Asians were only represented at the HFS-Other locations, and they use a slightly higher average than other races among the HFS-Other locations, except for Native Americans. When compared to all of the facilities, Caucasian clients require the least amount of service hours. Again, the most significant differences between clients seems to be based on facility type – with the biggest difference between schools and reporting centers, followed by schools and sanctuary houses.

**Conclusion from the above 3 graphs:** Regardless of demographics, a closer examination of the above three plots reveals that clients visiting the same institutions receive similar treatment hours. Clients at schools, for example, have the same treatment hours regardless of their demographics. This information may aid HFS in determining whether facilities require additional investment in order to reduce employee service hours and client treatment hours. This research also helps HFS determine which demographic client should be assigned to which member of staff or facility in order to improve program efficiency. 
