---
title: "R Script"
output: github_document
---

**Team 1 - Akhil Kodali, Heather Davis, Scott Bui**


# Data Cleaning

Same data cleaning we performed before with one slight modification identified in the "To modify the values containing "NA" to "Not Obtained"" and the "To select the columns that are required for our analysis" sections. 

## Loading the packages for select() and plots

```{r load package}
library(dplyr)
library(ggplot2)
```


## To load the data

```{r load data}
Full_HFS_Data <- read.csv("HFS Service Data.csv")
```
 

## To select the columns that are required for our analysis

```{r select columns}
HFS_1 <- select(Full_HFS_Data, c(program_name, facility, job_title,staff_name, actual_date, event_name, is_approved, NormalWorkHours, total_duration_num, recordID, state,age, simple_race,ethnic_identity, gender_identity))
```

This time we included the state and age columns. 


## To select the data that has is_approved = 1

```{r is_approved}
HFS_2 <- subset(HFS_1, is_approved == 1) 
```

Here is_approved = 1 means the people who are approved for the service. So, only approved data and patients who have visited the service are taken into consideration as there is no point in analyzing the data for which the patients have not been treated.
 
 
## To omit the data that has  simple_race = 21,65,5,9,17,20,81

```{r simple_race outliers}
HFS_Data <- subset(HFS_2,  simple_race != 21 & simple_race != 65 &simple_race != 5 & simple_race != 9 & simple_race != 17 & simple_race != 20  & simple_race != 81 )
```
 

## To replace the numeric codes in simple_race to actual words.

```{r simple_race labels}
HFS_Data$simple_race[HFS_Data$simple_race ==1 ] <- "Caucasian"
HFS_Data$simple_race[HFS_Data$simple_race ==2 ] <- "Alaskan Native"
HFS_Data$simple_race[HFS_Data$simple_race ==4 ] <- "American Indian/Native American"
HFS_Data$simple_race[HFS_Data$simple_race ==8 ] <- "Asian"
HFS_Data$simple_race[HFS_Data$simple_race ==16 ] <- "Black / African-American"
HFS_Data$simple_race[HFS_Data$simple_race ==32] <- "Hawaiian Native / Pacific Islander"
HFS_Data$simple_race[HFS_Data$simple_race ==64] <- "Two or more races"
HFS_Data$simple_race[HFS_Data$simple_race ==128 ] <- "Other"
HFS_Data$simple_race[HFS_Data$simple_race ==0 ] <- "Unknown or Not Collected"
```
 

## To modify the column gender_identity so that all the values having value "Female" are converted to "Woman"

```{r gender_identity female to woman}
HFS_Data$gender_identity[HFS_Data$gender_identity == "Female"] <- "Woman"
```
 
## To modify the values containing "NA" to "Not Obtained

```{r gender_identity Not Obtained to NA}
HFS_Data[is.na(HFS_Data)] <- "Not Obtained"
```


## To verify data has been modified according to our needs

```{r cleanup}
sort(unique(HFS_Data$simple_race))
unique(HFS_Data$gender_identity)
unique(HFS_Data$is_approved)
```
 

## To view the modified data

```{r view}
str(HFS_Data)
```

# Plots


## Hours and Facilities, Faceted by Gender Identity

```{r hours and facilities with gender id facets}
ggplot(data = HFS_Data) + 
geom_point(mapping = aes(x = facility, y = total_duration_num)) + 
    theme(axis.text.x = element_text(angle=90, vjust=1, hjust=1))+
facet_grid(. ~gender_identity)
```

In this visualization, we are observing whether we could identify some trends by faceting by gender identity. We were unable to see much in the facilities because there were too many variables. There is one pattern visible, where Two-Spirit appears to be at one facility. For the next steps on clarifying this visual, I would like to remove the upper end outliers in the Woman column so the dots are clearly visible. I would also like to aggregate the facilities to see better on what is happening across facility types and facility locations. 

## Hours and Gender with Facility Color Coding

```{r hours and gender id with facility bar}
ggplot(data = HFS_Data) + 
geom_col(mapping = aes(x = gender_identity, y = total_duration_num, color = facility))
```

In the normal view of this visualization, it can be observed that people who identify as women make up more of the total duration hours than any other gender identity. Like the previous plot, I would like to explore removing the top end outliers in the woman column and consolidate the facilities by type and location to see if it may yield more patterns. 

## Hours and Gender Identity Only

```{r hours and gender id bar}
ggplot(data = HFS_Data) + 
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))+
geom_col(mapping = aes(x = gender_identity, y = total_duration_num))
```

Like the previous plot, this bar chart indicates that people who identify as women make up more of the total duration hours than any other gender identity. There may not be a need to further modify this plot for observation. 

## Hours and Gender with Program Name Legend

```{r hours and gender with program name legend}
ggplot(data = HFS_Data) + 
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))+
geom_col(mapping = aes(x = gender_identity, y = total_duration_num, color = program_name) )
```

This bar chart indicates that most of the program provides services for mental health and substance use, with the majority of services provided being mental health. There isn't a need to drill this information any differently, as there are no further interesting patterns to discover. 

## Hours and Race faceted by different programs

```{r Hours and Race with Program Name Facet}
DurationRace <- ggplot(data=HFS_Data, aes(x=simple_race, y=total_duration_num))+
  facet_grid(.~program_name)+
geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle=60, vjust=, hjust=1))
options(scipen = 999)
DurationRace
```

From analysis, the group with the highest duration time utilized for the service of mental health and substance abuse are the Caucasian group. The second highest in both programs is Black/African-American. There is very limited help that is needed in regards of Gambling in all race groups.

## Hours and Age faceted by different programs

```{r Hours and Age with Program Name Facet}
DurationAge <- ggplot(data=HFS_Data, aes(x=age, y=total_duration_num))+
geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))+
 facet_grid(.~program_name)
options(scipen = 999)

DurationAge
```

Mental health seems to be the highest and most populated health concern for all age groups. We can take away that ages 7-10 has the highest spike utilizing help and services. Coming in second for the mental health category are years 28-30. Substance use has it's highest marks in the group's late 20's specifically ages 26-30.

## Count comparison of three program groups with different events

```{r}
testing2 <- ggplot(data=HFS_Data, aes(x = event_name)) +
  geom_bar(fill = "#112446") +
      theme(axis.text.x = element_text(angle=70, vjust=1, hjust=1))+
facet_grid(.~program_name)
testing2
```

This exploration attempt was to see which event was being utilized the most by which program. Gambling will be removed because there isn't much data in that category, so we'll need to adjust the plot to remove that. The mental health category will be the main focus. Substance abuse has some information, until variables are cleaned up there cannot be a solid observation with the graph alone. For the next iteration of this visualization, a cleaner graph will be used to tell a better story, by omitting the unneeded information.


## Box Plots of different programs w.r.t total duration

```{r}
ggplot(HFS_Data, aes(program_name,total_duration_num,by1=recordID,fill=program_name)) + 
  geom_boxplot(aes(program_name), alpha=0.3) +
  xlab("Programs offered") +ylab("Total Duration")
  theme(legend.position = "none")
```

The above graph shows that Gambling has a very low average compared with the Mental Health and Substance Use. Substance Use and Mental Health have close to the same average number of hours for treatment. There are some outliers in Mental Health, which explains why some patients are taking more time compared to other programs. From this observation, we can state that Gambling's distribution is skewed high right. Mental Health shows a highly left skewed distribution and Substance Use shows a normal distribution of data.


## To determine the total number of hours spent by various age groups in all states


```{r}
HFS_Data <- HFS_Data %>% mutate(agegroup = case_when(age >= 63  & age <= 72 ~ '63-72',
                                                     age >= 54  & age <= 63 ~ '54-63',
                                                     age >= 45  & age <= 54 ~ '45-54',
                                                     age >= 37  & age <= 45 ~ '37-45',
                                                     age >= 28  & age <= 36 ~ '28-36',
                                                     age >= 19  & age <= 27 ~ '19-27',
                                                     age >= 10  & age <= 18 ~ '10-18',
                                                     age >= 0  & age <= 8 ~ '1-9'))



ggplot(data = HFS_Data) +   
  
  geom_bar(stat = "identity",mapping = aes(x = agegroup, y = total_duration_num/60, fill =state)) +   
  
  facet_wrap( ~program_name,ncol=1) + labs(y= "Total duration in hours", x = "Different Age Groups")
```

This bar chart appears interesting when we compare Iowa and Nebraska with Mental Health and Gambling. The people in Iowa are taking more time to normalize for Substance Use and the people in Nebraska are taking more time for Mental Health. Next time, this plot can be drawn using another parameter record_id, as this may help enable return of actual information based on individuals, rather than comparing all. 


## Total visits by years from the start of service for different program names by different facilities

```{r warning=FALSE}
HFS_Data$actual_date <- as.numeric(HFS_Data$actual_date)
HFS_Data <- HFS_Data %>% mutate(actual_date = case_when(actual_date >= 3287  & actual_date <= 3652 ~ 'Year10',
                                                        actual_date >= 2923  & actual_date <= 3287 ~ 'Year9',
                                                        actual_date >= 2557  & actual_date <= 2922 ~ 'Year8',
                                                        actual_date >= 2192  & actual_date <= 2556 ~ 'Year7',
                                                        actual_date >= 1827  & actual_date <= 2191 ~ 'Year6',
                                                        actual_date >= 1462  & actual_date <= 1826 ~ 'Year5',
                                                        actual_date >= 1096  & actual_date <= 1461 ~ 'Year4',
                                                        actual_date >= 731  & actual_date <= 1095 ~ 'Year3',
                                                        actual_date >= 366  & actual_date <= 730 ~ 'Year2',
                                                        actual_date >= -7  & actual_date <= 365   ~ 'Year1'))



nofvisits <- data.frame(count(HFS_Data,actual_date,program_name,facility))
ggplot(data = nofvisits,aes(x=actual_date,y=n,fill=facility,na.rm=TRUE))+geom_bar(stat="identity")+facet_wrap(~program_name) 
```

In this plot, we can see that the total number of visits are decreasing, which can be considered a good sign as the visits are decreasing year by year. In the next task for the extension of this, we can find out if it is the people or the visits that are decreasing using record_id. 

## Number of staff over the years

```{r}
da<-HFS_Data %>%
  count(actual_date,staff_name,job_title)

da2<-as.data.frame(table(da$actual_date,da$job_title))

ggplot(data=da2, aes(x=Var1,y=Freq))+geom_bar(stat="identity")+labs(y= "Total Resources", x = "Age Groups")
```

From the above plot we can observe that the working staff also kept on decreasing for years through which we can think that HFS is working more towards saving their funds.


# Contributor Statement

Heather Davis, Scott Bui, and Akhil Kodali each created plots. Heather Davis wrote the first draft of the markdown document and Scott Bui added the remaining code and plots to it. Heather Davis proofed the document, updated the ReadMe with a link, and submitted it for the assignment.
